Concurrent Programming in Harmony

### 2021-12-26 book/paper.tex
Working on formatting

<{:const:f}>
<{:const:g}>

@@

Modern databases support multiple clients concurrently accessing the data.
They store data on disk, but we will ignore that in this book.
(If you want to model a disk, this is probably best done as a separate
thread.)
The complication we address here is that databases may be \emph{sharded},
where different parts of the data are stored on different servers.
The different servers may even be under different
authoritive domains, such as multiple banks maintaining the accounts of their
clients.

In database terminology, a \emph{transaction} is an operation on a database.
The operation can be quite complex, and the execution of a transaction
should have at least the following two properties:
\begin{itemize}
\item \emph{all-or-nothing}: a transaction should either complete, or it should
be a no-op.  It should never partially execute and then give up because of
some kind of error or something.  Database
people call this \emph{atomicity}, but it is not the same kind of atomicity that
we have been discussing in this book.
\item \emph{serialized}: any two concurrent transactions should appear to
execute in some order.  Database people call this \emph{isolation}: one transaction
should not be able to witness the intermediate state of another transaction
in execution.
\end{itemize}

We will use as an example a distributed database that maintains accounts.
For simplicity we will model this as a collection of banks, each maintaining a
single account.  There are two kinds of transactions: \emph{transfer}
(similar to \autoref{ex:bank}) and \emph{check}.  In this example, a transfer
is a transaction that moves some funds between two accounts.
A check is a transaction over all accounts and checks that the sum of the
balances across the accounts remains the same.

@@

The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
used to implement transactions across multiple database servers---banks
in this case.
Each transaction has a \emph{coordinator} that sends a <{PREPARE}>
message to each of the servers involved in the transaction, asking them
to prepare to commit to their part in a particular transaction.
A server can either respond with
\texttt{YES} if it is ready to commit and will avoid doing anything
that might jeopardize this (like committing a conflicting transaction),
or with \texttt{NO} if it does not want to participate in the transaction.
If all servers respond with \texttt{YES}, then the coordinator can
decide to \emph{commit} the transaction.  Otherwise the coordinator
must decide to \emph{abort} the transaction.  In the second phase, the
servers that responded with \texttt{YES} (if any) must be notified
to inform them of the coordinator's decision.

@@

\begin{itemize}
\item <{False < True}>;
\item integers are ordered in the natural way;
\item strings are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of Harmony values.  <{None}> is the empty list.
Addresses are lexicographically ordered accordingly;
\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content.
\end{itemize}

@@

The boolean type has only two possible values: <{False}> and
<{True}>.  Unlike in Python, in Harmony booleans are distinct
from integers, and in particular $\mathbf{False} < 0$.  In statements
and expressions where booleans are expected, it is not possible to
substibute values of other types.

@@

A string is a sequence of zero or more unicode characters.
If they do not contain special characters and do not start with a digit,
then a string can be represented by a ``.'' followed by the characters.
For example, <{.example}> is the same as the string <{"example"}>.

@@

A dictionary maps a set of values (known as \emph{keys})
to another set of values.
The generic syntax of a dictionary is
$\{ k_0:v_0, k_1:v_1, ... \}$.
Different from Python, the empty dictionary is either <{()}> or $[]$
(because $\{\}$ is the empty set in Harmony).
If there are duplicate keys in the list, then only the one with
the maximum value survives.
Therefore the order of the keys in the dictionary does not matter.

Dictionaries support comprehension.  The basic form is:
\texttt{\{ f(<{k}>):g(<{k}>) for <{k}> in <{s}> \}}.

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{d k}> & indexing \\
\hline
<{len d}> & the number of keys in <{d}> \\
\hline
<{keys d}> & the set of keys in <{d}> \\
\hline
<{<{v [not] in keys d}> & check if <{v}> is [not] a key in <{d}> \\
\hline
<{v [not] in d}> & check if <{v}> is [not] a value in <{d}> \\
\hline
<{min d}> & the minimum value in <{d}> \\
\hline
<{max d}> & the maximum value in <{d}> \\
\hline
<{any d}> & <{True}> if any value is <{True}> \\
\hline
<{all d}> & <{True}> if all values are <{True}> \\
\hline
<{d & d & ...}> & dictionary intersection \\
\hline
<{d | d | ...}> & dictionary union \\
\hline
\end{tabular}
\end{center}

Because in Harmony brackets are used for parsing purposes only,
you can write $d[k]$ (or $d(k)$) instead of $d~k$.
However, if <{k}> is a string like <{.id}>, then you might
prefer the notation $k\mathtt{.id}$.

Python users beware: the Harmony \texttt{<{v}> in <{d}>} operator checks
if there is some key <{k}> such that $d[k] = v$.  In Python, the same
syntax checks if <{v}> is a key in <{d}>.  The difference exists because
in Harmony a list is a special case of a dictionary.

@@

Lists and tuples support comprehension as well.
In their most basic form: \texttt{f(<{v}>) for <{v}> in <{s}>}.
For example, to check if any element in a list <{t}> is even, you
can write:

\texttt{any($(x \% 2) == 0$ for <{x}> in <{t}>)}.

@@

You can create lambda functions similarly to Python.  For example:
\texttt{lambda$(x,y)$:$x+y$}.

@@

A thread can call \texttt{get\_context()} to retrieve its current
context.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{e}> & <{e}> is an expression\\
\hline
<{lv = [lv =]... e}> & <{lv}> is an lvalue and <{e}> is an expression\\
\hline
<{lv} [op]= e}> & <{op}> is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^}, <{and}>, <{or}>\\
\hline
<{l: S}> & <{l}> is a label \\
\hline
<{assert b [, e]}> & <{b}> is a boolean.  Optionally report value of expression <{e}>\\
\hline
<{await b}> & <{b}> is a boolean\\
\hline
<{const a = e}> & <{a}> is a bound variable, <{e}> is a constant expression\\
\hline
<{def m a: S}> & <{m}> is an identifier, <{a}> a bound variable\\
\hline
<{del lv}> & delete <{lv}>\\
\hline
<{for a[:b] in e [where c]: S}> & <{a}> and <{b}> are bound variables, <{e}> is a set or dictionary \\
\hline
<{from m import ...}> & <{m}> identifies a module \\
\hline
<{go c e}> & <{c}> is a context, <{e}> is an expression \\
\hline
<{if b: S else: S}> & <{b}> is a boolean, <{S}> is a list of statements\\
\hline
<{import m, ...}> & <{m}> identifies a module \\
\hline
<{invariant e}> & <{e}> is an invariant\\
\hline
<{let a = e: S} & <{a}> is a bound variable, <{e}> is an expression\\
\hline
<{pass}> & do nothing\\
\hline
<{print e>}> & <{e}> is an expression \\
\hline
<{sequential v, ...}> & <{v}> has sequential consistency \\
\hline
<{spawn [eternal] m e [, t]}> & <{m}> is a method,
<{e}> is an expression, <{t}> is the thread-local state \\
\hline
<{trap m e}> & <{m}> is a method and <{e}> is an expression \\
\hline
<{var v = e}> & <{v}> is a new variable, <{e}> is an expression\\
\hline
<{when b: S}> & <{b}> is a boolean, <{S}> a list of statements\\
\hline
<{when exists a in e: S}> & <{a}> is a bound variable, <{e}> is an expression\\
\hline
<{while b: S}> & <{b}> is a boolean, <{S}> a list of statements\\
\hline
\end{tabular}

@@

The statement <{x = 3}> changes the state by assigning~3 to
variable <{x}> (assuming~<{x}> was not already~3).
<{x}> may be a local variable or a shared variable.
The statement $x = y = 3$ first updates~<{y}>, then~<{x}>.
The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = \mathtt{h}()$ first computes
the address of $x[\mathtt{f}()]$, then computes the address of
$y[\mathtt{g}()]$, then evaluates $\mathtt{h}()$, then assigns the resulting value to $y[\mathtt{g}()]$ (using its previously computed address), and finally assigns the same value to $x[\mathtt{f}()]$ (again using its previously computed address).
The statement $a,b = c$ assumes that <{c}> is a tuple with two values.
It first evaluates the addresses of <{a}> and <{b}> and first assigns to
the latter and then the former.
If <{c}> is not a tuple with two values, then Harmony will report an error.

Assigning to $\_$ evaluates the righthand side expression but is otherwise
a no-op.

The statement $x~+$$= 3$ loads <{x}>, adds 3, and then stores the
results in <{x}>.  In this case, it is equivalent to $x = x + 3$.
However, in general this is not so.  For example, $x[\mathtt{f}()]~+$$= 3$
only evaluates $\mathtt{f}()$ once.  Unlike Python, however,
$x~+$$= [3,]$ is equivalent to $x = x + [3,]$ in Harmony.
(In Python, the following two compound statements lead to different results
for <{y}>: $x = y = []; x~+$$= [3]$ and $x = y = []; x = x + [3]$.)

>>>>

<{:const:f}>
<{:const:g}>
<{:const:h}>

@@

Modern databases support multiple clients concurrently accessing the data.
They store data on disk, but we will ignore that in this book.
(If you want to model a disk, this is probably best done as a separate
thread that maintains the contents of the disk.)
The complication we address here is that databases may be \emph{sharded},
where different parts of the data are stored on different servers.
The different servers may even be under different
authoritive domains, such as multiple banks maintaining the accounts of their
clients.

In database terminology, a \emph{transaction} is an operation on a database.
The operation can be quite complex, and the execution of a transaction
should have the following two properties:
\begin{itemize}
\item \emph{all-or-nothing}: a transaction should either complete, or it should
be a no-op.  It should never partially execute and then give up because of
some kind of error or something.  Database
people call this \emph{atomicity}, but it is not the same kind of atomicity that
we have been discussing in this book.
\item \emph{serialized}: any two concurrent transactions should appear to
execute in some order.  Database people call this \emph{isolation}: one transaction
should not be able to witness the intermediate state of another transaction
in execution.
\end{itemize}

We will use as an example a distributed database that maintains bank accounts.
For simplicity, we will model this as a collection of banks,
each maintaining a single account.
There are two kinds of transactions: \emph{transfer}
(similar to \autoref{ex:bank}) and \emph{check}.  In this example, a transfer
is a transaction that moves some funds between two accounts.
A check is a transaction over all accounts and checks that the sum of the
balances across the accounts remains the same.

@@

The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
used to implement transactions across multiple database servers---banks
in this case.
Each transaction has a \emph{coordinator} that sends a <{PREPARE}>
message to each of the servers involved in the transaction, asking them
to prepare to commit to their part in a particular transaction.
A server can either respond with
<{YES}> if it is ready to commit and will avoid doing anything
that might jeopardize this (like committing a conflicting transaction),
or with <{NO}> if it does not want to participate in the transaction.
If all servers respond with <{YES}>, then the coordinator can
decide to \emph{commit} the transaction.  Otherwise the coordinator
must decide to \emph{abort} the transaction.  In the second phase, the
servers that responded with <{YES}> (if any) must be notified
to inform them of the coordinator's decision.

@@

\begin{itemize}
\item <{False < True}>;
\item integers are ordered in the natural way;
\item strings are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of Harmony values.  <{None}> is the empty list.
Addresses are lexicographically ordered accordingly;
\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content in an unspecified way.
\end{itemize}

@@

The boolean type has only two possible values: <{False}> and
<{True}>.  Unlike in Python, in Harmony booleans are distinct
from integers, and in particular $\mathbf{False} < 0$.  In statements
and expressions where booleans are expected, it is not possible to
substitute values of other types.

@@

A string is a sequence of zero or more unicode characters.
If it consists entirely of alphanumerical characters or underscore
characters and does not start with a digit,
then a string can be represented by a ``.'' followed by the characters.
For example, <{.example}> is the same as the string <{"example"}>.

@@

A dictionary maps a set of values (known as \emph{keys})
to another set of values.
The generic syntax of a dictionary is
$\{ k_0:v_0, k_1:v_1, ... \}$.
Different from Python, the empty dictionary is either <{()}> or <{[]}>
(because <{{}}> is the empty set in Harmony).
If there are duplicate keys in the list, then only the one with
the maximum value survives.
Therefore the order of the keys in the dictionary does not matter.

Dictionaries support comprehension.  The basic form is:
<{{ f(k):g(k) for k in s }}>.

@@

\begin{center}
\begin{tabular}{|l|l|}
\hline
<{d k}> & indexing \\
\hline
<{len d}> & the number of keys in <{d}> \\
\hline
<{keys d}> & the set of keys in <{d}> \\
\hline
<{v [not] in keys d}> & check if <{v}> is [not] a key in <{d}> \\
\hline
<{v [not] in d}> & check if <{v}> is [not] a value in <{d}> \\
\hline
<{min d}> & the minimum value in <{d}> \\
\hline
<{max d}> & the maximum value in <{d}> \\
\hline
<{any d}> & <{True}> if any value is <{True}> \\
\hline
<{all d}> & <{True}> if all values are <{True}> \\
\hline
<{d & d & ...}> & dictionary intersection \\
\hline
<{d | d | ...}> & dictionary union \\
\hline
\end{tabular}
\end{center}

Because in Harmony brackets are used for parsing purposes only,
you can write <{d[k]}> (or <{d(k)}>) instead of <{d k}>.
However, if <{k}> is a string like <{.id}>, then you might
prefer the notation <{k.id}>.

Python users beware: the Harmony <{v in d}> operator checks
if there is some key <{k}> such that <{d[k] = v}>.  In Python, the same
syntax checks if <{v}> is a key in <{d}>.  The difference exists because
in Harmony a list is a special case of a dictionary.

@@

Lists and tuples support comprehension as well.
In their most basic form: <{(f(v) for v in s)}>.
For example, to check if any element in a list <{t}> is even, you
can write: <{any((x % 2) == 0 for x in t)}>.

@@

You can create lambda functions similarly to Python, except that the
expression has to end on the keyword <{end}>.  For example:
<{lambda(x,y): x+y end}>.

@@

A thread can call <{get_context()}> to retrieve its current
context.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{e}> & <{e}> is an expression\\
\hline
<{lv = [lv =]... e}> & <{lv}> is an lvalue and <{e}> is an expression\\
\hline
<{lv} [op]= e}> & <{op}> is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^}, <{and}>, <{or}>\\
\hline
<{l: S}> & <{l}> is a label \\
\hline
<{assert b [, e]}> & <{b}> is a boolean.  Optionally report value of expression <{e}>\\
\hline
<{await b}> & <{b}> is a boolean\\
\hline
<{const a = e}> & <{a}> is a bound variable, <{e}> is a constant expression\\
\hline
<{def m a: S}> & <{m}> is an identifier, <{a}> a bound variable\\
\hline
<{del lv}> & delete <{lv}>\\
\hline
<{for a[:b] in e [where c]: S}> & <{a}> and <{b}> are bound variables, <{e}> is a set or dictionary \\
\hline
<{from m import ...}> & <{m}> identifies a module \\
\hline
<{go c e}> & <{c}> is a context, <{e}> is an expression \\
\hline
<{if b: S else: S}> & <{b}> is a boolean, <{S}> is a list of statements\\
\hline
<{import m, ...}> & <{m}> identifies a module \\
\hline
<{invariant e}> & <{e}> is an invariant\\
\hline
<{let a = e: S}> & <{a}> is a bound variable, <{e}> is an expression\\
\hline
<{pass}> & do nothing\\
\hline
<{print e>}> & <{e}> is an expression \\
\hline
<{sequential v, ...}> & <{v}> has sequential consistency \\
\hline
<{spawn [eternal] m e [, t]}> & <{m}> is a method,
<{e}> is an expression, <{t}> is the thread-local state \\
\hline
<{trap m e}> & <{m}> is a method and <{e}> is an expression \\
\hline
<{var v = e}> & <{v}> is a new variable, <{e}> is an expression\\
\hline
<{when b: S}> & <{b}> is a boolean, <{S}> a list of statements\\
\hline
<{when exists a in e: S}> & <{a}> is a bound variable, <{e}> is an expression\\
\hline
<{while b: S}> & <{b}> is a boolean, <{S}> a list of statements\\
\hline
\end{tabular}

@@

The statement <{x = 3}> changes the state by assigning~3 to
variable <{x}> (assuming~<{x}> was not already~3).
<{x}> may be a local variable or a shared variable.
The statement <{x = y = 3}> first updates~<{y}>, then~<{x}>.
The statement <{x[f()] = y[g()] = h()}> first computes
the address of <{x[f()]}>, then computes the address of
<{y[g()]}>, then evaluates <{h()}>,
then assigns the resulting value to <{y[g()]}>
(using its previously computed address),
and finally assigns the same value to <{x[f()]}>
(again using its previously computed address).
The statement <{a,b = c}> assumes that <{c}> is a tuple with two values.
It first evaluates the addresses of <{a}> and <{b}> and first assigns to
the latter and then the former.
If <{c}> is not a tuple with two values, then Harmony will report an error.

Assigning to <{_}> evaluates the righthand side expression but is otherwise
a no-op.

The statement $x~+$$= 3$ loads <{x}>, adds 3, and then stores the
results in <{x}>.  In this case, it is equivalent to $x = x + 3$.
However, in general this is not so.  For example, <{x[f()] += 3}>
only evaluates <{f()}> once.  Unlike Python, however,
<{x += [3,]}> is equivalent to <{x = x + [3,]}> in Harmony.
(In Python, the following two compound statements lead to different results
for <{y}>: <{x = y = []; x += [3]}> and <{x = y = []; x = x + [3]}>.)
